---
title: "Decathlon"
format: html
editor: visual
---

# Background Info

Decathlon is a two-day event that consists of ten track and field events. On the first day, athletes compete in the 100m, Long Jump, Shot Put, High Jump, and 400m. On the second day, athletes compete in the 110m Hurdles, Discuss, Pole Vault, Javelin Throw, and 1500m. 

Athletes amass points based off their individual performance marks rather than relative placement. Points are calculated using specific formulas for the track and field events, each having their own unique [constants](https://sport.maths.org/content/decathlon-art-scoring-points-0). 

```{r Loading Libraries}

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(mclust)
library(gt)
library(broom)
library(sportyR)

```

This dataset contains 12,762 elite men's decathlon performances from 2001 to 2022. Elite is defined as an overall score of 6400 or greater. 

```{r Loading Data}

dec_performances <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/dec_performances.csv")

```

In order to further explore the data, the variables: `day1_score`, `day2_score`, `exact_age`, `age`, `elite_performance`, and `pr` were added to the dataset. 

```{r Adding Variables}

dec_performances <- dec_performances |>
  mutate(day1_score = men_100_score + men_lj_score + men_sp_score + men_hj_score + men_400_score,
         day2_score = men_110h_score + ifelse(is.na(men_dt_score), 0, men_dt_score) + men_pv_score + men_jt_score + men_1500_score)

dec_performances <- dec_performances |>
  mutate(dob = as.Date(dob, format = "%m/%d/%Y"), 
         date = as.Date(date, format = "%m/%d/%Y"), 
         exact_age = interval(dob, date) / years(1), 
         exact_age = ifelse(exact_age > 95, NA, exact_age), 
         age = floor(exact_age))
 
dec_performances <- dec_performances |>
  group_by(competitor) |>
  arrange(date) |>
  mutate(elite_performance = row_number(), 
          pr = max(overall_score)) |>
  ungroup()

```

In order to explore questions related to scores, a data frame including the constants associated with each event was created. Additional variables were created to discover what necessary marks would to be met to get a certain number of points and what the difference in such marks are. 

```{r}

score_calculations <- data.frame("Event" = c("100m", "Long Jump", "Shot Put", "High Jump",
                                             "400m", "110m Hurdles", "Discuss Throw", 
                                             "Pole Vault", "Javelin Throw", "1500m"), 
                            "A" = c(25.4347, 0.14354, 51.39, 0.8465, 1.53775, 
                                    5.74352, 12.91, 0.2797, 10.14, 0.03768), 
                            "B" = c(18, 220, 1.5, 75, 82, 28.5, 4, 100, 7, 480), 
                            "C" = c(1.81, 1.4, 1.05, 1.42, 1.81, 1.92, 1.1, 1.35, 1.08, 1.85))

score_calculations <- score_calculations |>
  mutate(get_500_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (500/A)^(1/C), 
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (500/A)^(1/C)) / 100, B + (500/A)^(1/C))),
         get_600_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"),
                                 B - (600/A)^(1/C),
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (600/A)^(1/C)) / 100, B + (600/A)^(1/C))), 
         get_700_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (700/A)^(1/C), 
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (700/A)^(1/C)) / 100, B + (700/A)^(1/C))), 
         get_800_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (800/A)^(1/C), 
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (800/A)^(1/C)) / 100, B + (800/A)^(1/C))),
         get_900_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (900/A)^(1/C),
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (900/A)^(1/C)) / 100, B + (900/A)^(1/C))), 
         get_1000_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (1000/A)^(1/C),
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (1000/A)^(1/C)) / 100, B + (1000/A)^(1/C))),
         get_1100_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (1100/A)^(1/C),
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (1100/A)^(1/C)) / 100, B + (1100/A)^(1/C))),
         get_1200_points = ifelse(Event %in% c("100m", "400m", "110m Hurdles", "1500m"), 
                                 B - (1200/A)^(1/C),
                                 ifelse(Event %in% c("Long Jump", "High Jump", "Pole Vault"),
                                        (B + (1200/A)^(1/C)) / 100, B + (1200/A)^(1/C))),
         increase1 = (get_600_points - get_500_points) / 1,
         increase2 = (get_700_points - get_600_points) / 1,
         increase3 = (get_800_points - get_700_points) / 1,
         increase4 = (get_900_points - get_800_points) / 1,
         increase5 = (get_1000_points - get_900_points) / 1,
         increase6 = (get_1100_points - get_1000_points) / 1,
         increase7 = (get_1200_points - get_1100_points) / 1)

```

In order to investigate how the difficulty of improving in each event, we will create a dataset of the athletes with their highest and lowest scores in each event and find the difference between them. Here, only competitors with 5 or more elite performances were included. 

```{r}

event_improvements <- dec_performances |> 
  group_by(competitor) |>
  filter(n() >= 5) |>
  summarize(low_100m = min(men_100_score), 
            high_100m = max(men_100_score), 
            low_lj = min(men_lj_score), 
            high_lj = max(men_lj_score), 
            low_sp = min(men_sp_score), 
            high_sp = max(men_sp_score), 
            low_hj = min(men_hj_score), 
            high_hj = max(men_hj_score), 
            low_400m = min(men_400_score), 
            high_400m = max(men_400_score), 
            low_110h = min(men_110h_score), 
            high_110h = max(men_110h_score), 
            low_dt = min(men_dt_score), 
            high_dt = max(men_dt_score), 
            low_pv = min(men_pv_score), 
            high_pv = max(men_pv_score), 
            low_jt = min(men_jt_score), 
            high_jt = max(men_jt_score), 
            low_1500m = min(men_1500_score), 
            high_1500m = max(men_1500_score), 
            low_overall = min(overall_score),
            high_overall = max(overall_score)) |>
  ungroup()

event_improvements <- event_improvements |>
  mutate(improvement_100m = high_100m - low_100m,
         improvement_lj = high_lj - low_lj, 
         improvement_sp = high_sp - low_sp, 
         improvement_hj = high_hj - low_hj, 
         improvement_400m = high_400m - low_400m, 
         improvement_110h = high_110h - low_110h,
         improvement_dt = high_dt - low_dt, 
         improvement_pv = high_pv - low_pv, 
         improvement_jt = high_jt - low_jt, 
         improvement_1500m = high_1500m - low_1500m, 
         improvement_overall = high_overall - low_overall)

```

# Summaries

```{r Event Scores}

summary(dec_performances$men_100_score)
summary(dec_performances$men_lj_score)
summary(dec_performances$men_sp_score)
summary(dec_performances$men_hj_score)
summary(dec_performances$men_400_score)
summary(dec_performances$men_110h_score)
summary(dec_performances$men_dt_score)
summary(dec_performances$men_pv_score)
summary(dec_performances$men_jt_score)
summary(dec_performances$men_1500_score)

```

```{r}

summary(dec_performances$day1_score)
summary(dec_performances$day2_score)
summary(dec_performances$overall_score)
summary(dec_performances$world_athletics_event_ranking_score)

t.test(dec_performances$day1_score, dec_performances$day2_score, paired = TRUE)
```

An interesting finding here is that there is a statistically significant difference in the means of decathlon day 1 scores and decathlon day 2 scores.  

```{r Athletes by PR}

dec_performances |>
  group_by(competitor) |>
  summarize(pr = max(pr)) |>
  arrange(desc(pr))

```

```{r}

summary(event_improvements$improvement_100m)
summary(event_improvements$improvement_lj)
summary(event_improvements$improvement_sp)
summary(event_improvements$improvement_hj)
summary(event_improvements$improvement_400m)
summary(event_improvements$improvement_110h)
summary(event_improvements$improvement_dt)
summary(event_improvements$improvement_pv)
summary(event_improvements$improvement_jt)
summary(event_improvements$improvement_1500m)
summary(event_improvements$improvement_overall)

```

```{r}

event_improvements |> 
  summarize(average_100m = mean(improvement_100m), 
            average_lj = mean(improvement_lj), 
            average_sp = mean(improvement_sp), 
            average_hj = mean(improvement_hj), 
            average_400m = mean(improvement_400m), 
            average_110h = mean(improvement_110h), 
            average_dt = mean(improvement_dt, na.rm = TRUE), 
            average_pv = mean(improvement_pv), 
            average_jt = mean(improvement_jt), 
            average_1500m = mean(improvement_1500m), 
            average_overall = mean(improvement_overall)) |>
  pivot_longer(average_100m:average_overall, names_to = "Averages", values_to = "Value") |>
  arrange(desc(Value))

```

```{r}
event_improvements |> 
  summarize(average_100m = median(improvement_100m), 
            average_lj = median(improvement_lj), 
            average_sp = median(improvement_sp), 
            average_hj = median(improvement_hj), 
            average_400m = median(improvement_400m), 
            average_110h = median(improvement_110h), 
            average_dt = median(improvement_dt, na.rm = TRUE), 
            average_pv = median(improvement_pv), 
            average_jt = median(improvement_jt), 
            average_1500m = median(improvement_1500m), 
            average_overall = median(improvement_overall)) |>
  pivot_longer(average_100m:average_overall, names_to = "Averages", values_to = "Value") |>
  arrange(desc(Value))

```

# Visualizations

```{r}

dec_performances |>
  select(rank, overall_score, day1_score, day2_score) |>
  filter(rank < 100) |>
  pivot_longer(overall_score:day2_score, names_to = "score", values_to = "value") |>
  ggplot(aes(rank, value)) +
  geom_point(alpha = 0.2, color = "burlywood") + 
  facet_wrap(~ score, scales = "free_y") +
  labs(title = "Top 100 Scores by Rank", y = "Score", x = "Rank") +
  theme_bw() +
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5))

```

```{r}

dec_performances |>
  filter(rank <= 100) |>
  group_by(rank) |>
  summarize(average_overall_score = sum(overall_score) / n(), 
            average_day1_score = sum(day1_score) / n(),
            average_day2_score = sum(day2_score) / n()) |>
  pivot_longer(average_overall_score:average_day2_score, 
               names_to = "score", values_to = "value") |>
  ggplot(aes(rank, value)) +
  geom_point(alpha = 0.2, color = "burlywood") + 
  facet_wrap(~ score, scales = "free_y") +
  labs(title = "Top 100 Average Scores by Rank", y = "Score", x = "Rank") +
  theme_bw() +
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5))

```

```{r}

dec_performances |>
  summarize(average_100_score = sum(men_100_score) / n(),
            average_lj_score = sum(men_lj_score) / n(),
            average_sp_score = sum(men_sp_score) / n(),
            average_hj_score = sum(men_hj_score) / n(),
            average_400_score = sum(men_400_score) / n(),
            average_110h_score = sum(men_110h_score) / n(),
            average_dt_score = sum(men_dt_score, na.rm = TRUE) / (n() - sum(is.na(men_dt_score))),
            average_pv_score = sum(men_pv_score) / n(),
            average_jt_score = sum(men_jt_score) / n(),
            average_1500_score = sum(men_1500_score) / n()) |>
  pivot_longer(average_100_score:average_1500_score, 
               names_to = "Event", values_to = "Score") |>
  ggplot(aes(x = fct_reorder(Event, Score, .desc = TRUE), y = Score)) +
  geom_col(fill = "cornflowerblue") + 
  theme_bw() +
  labs(title = "Average Event Score", y = "Average Score", x = "Event") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
        plot.title = element_text(hjust = 0.5))

```

```{r}

dec_performances |>
  select(competitor, overall_score, elite_performance, pr) |>
  filter(pr >= 8670) |>
  ggplot(aes(x = elite_performance, y = overall_score)) +
  geom_line() + 
  geom_point(color = "tomato", alpha = 0.5) +
  facet_wrap(~ competitor) +
  labs(title = "Overall Score by Elite Performance", subtitle = "Top 16 Athletes by PR",
       y = "Overall Score", x = "Elite Performance") +
  theme_bw() +
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5),  
        plot.subtitle = element_text(hjust = 0.5))

```

```{r}

dec_performances |>
  select(competitor, overall_score, date, pr) |>
  filter(pr >= 8670) |>
  ggplot(aes(x = date, y = overall_score)) +
  geom_line() + 
  geom_point(color = "tomato", alpha = 0.5) +
  facet_wrap(~ competitor) + 
  labs(title = "Overall Scores by Date", subtitle = "Top 16 Athletes by PR", 
       y = "Overall Score", x = "Date") +
  theme_bw() +
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5), 
        plot.subtitle = element_text(hjust = 0.5))

```

```{r}

dec_performances |>
  select(men_100_score, men_110h_score, men_lj_score, men_dt_score, men_sp_score,
         men_pv_score, men_hj_score, men_jt_score, men_400_score, men_1500_score) |>
  pivot_longer(men_100_score:men_1500_score, 
               names_to = "Event", values_to = "Value") |>
  ggplot(aes(Value)) +
  geom_boxplot() +
  facet_wrap(~ fct_inorder(Event), nrow = 5) + 
  theme_bw() + 
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5)) + 
  labs(title = "Event Score Summaries")

```

```{r}

score_calculations |>
  group_by(Event) |>
  pivot_longer(increase1:increase7, 
               names_to = "Increase", values_to = "Value") |>
  ggplot(aes(Increase, Value)) +
  geom_point() +
  facet_wrap(~ Event, scales = "free_y", nrow = 2) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
        strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5), 
        plot.subtitle = element_text(hjust = 0.5)) + 
  labs(title = "Needed Mark for 100 Point Increase by Event", 
       subtitle = "Intervals Cover 100 Point Increases from 500 Points to 1200 Points")

```

```{r}

event_improvements |>
  select(improvement_100m, improvement_110h, improvement_lj, improvement_dt, 
         improvement_sp, improvement_pv, improvement_hj, improvement_jt, 
         improvement_400m, improvement_1500m) |>
  pivot_longer(improvement_100m:improvement_1500m, 
               names_to = "Improvements", values_to = "Value") |>
  ggplot(aes(Value)) + 
  geom_boxplot() + 
  facet_wrap(~ fct_inorder(Improvements), ncol = 2) +
  theme_bw() +
  theme(strip.background = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Event Improvement Summaries")

```

```{r}

dec_performances |>
  filter(overall_score == 9126) |>
  select(men_100_score, men_lj_score, men_sp_score, men_hj_score, men_400_score,
         men_110h_score, men_dt_score, men_pv_score, men_jt_score, men_1500_score) |>
  pivot_longer(men_100_score:men_1500_score, names_to = "Scores", values_to = "Value") |>
  ggplot(aes(y = Value, x = fct_reorder(Scores, Value, .desc = TRUE))) +
  geom_col(fill = "blue2") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Kevin Mayer's WR Event Scores", y = "Score", x = "Event")
  
```

```{r Gaussian Mixture Model}

dec_mclust <- dec_performances |>
  select(men_100_score, men_lj_score, men_sp_score, men_hj_score, men_400_score,
         men_110h_score, men_dt_score, men_pv_score, men_jt_score, men_1500_score) |>
  mutate(men_dt_score = ifelse(is.na(men_dt_score), 0, men_dt_score)) |>
  Mclust()

```

```{r}

summary(dec_mclust)

```

```{r}

gt(tidy(dec_mclust))

dec_mclust |> 
  augment() |> 
  ggplot(aes(x = men_1500_score , y = men_400_score, 
             color = .class, size = .uncertainty)) +
  geom_point(alpha = 0.5)

dec_mclust |> 
  plot(what = "BIC", 
       legendArgs = list(x = "bottomright", ncol = 4))

dec_mclust |> 
  plot(what = "classification")

```

```{r}

table("Clusters" = dec_mclust$classification, "Age" = dec_performances$age)

```

```{r}

dec_probs <- dec_mclust$z
colnames(dec_probs) <- str_c("cluster", 1:8)

dec_probs <- dec_probs |>
  as_tibble() |>
  mutate(competitor = dec_performances$competitor, age = dec_performances$age) |>
  pivot_longer(-c(competitor, age), names_to = "cluster", values_to = "prob")

dec_probs |>
  ggplot(aes(prob)) +
  geom_histogram() +
  facet_wrap(~cluster)

```
